---
title: "snp-gene-assoc"
output: html_document
date: "2026-06-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  warning = FALSE,
  message = FALSE
)
```
------------------------------------------------------------------------

# Overview

This SOP documents the full analysis pipeline for candidate gene SNP
association analysis in livestock using a Fluidigm/Agena targeted SNP
panel. The pipeline covers genotype quality control, genetic diversity
estimation, population structure analysis, and SNP-trait association
analysis.

**Study design:**

| Item | Description |
|------------------------------------|------------------------------------|
| Animals | n = 89–92 (varies by available phenotype) |
| Breeds | KK, KK Elite (KKE), Nellore |
| Locations | PTH Jelai NS, PTH Tersat Terengganu, PTH Ulu Lepar Pahang, MARDI Kemaman Terengganu |
| Genotyping platform | Fluidigm 96.96 Dynamic Array |
| SNPs | 33 high-quality polymorphic SNPs (post-QC) |
| Traits | BW0, WW205, BW12, ADG_0-6, ADG_6-12, RGR |

------------------------------------------------------------------------

# Pipeline Flowchart

```text
Raw genotype data (HapMap .txt)
  ┌───────────────┐
  │   snpReady    │  → MAF filter & QC
  │               │  → Encode genotype to 0/1/2
  └───────┬───────┘  → HWE, Ho, He, MAF, PIC
          │
          ▼
  ┌───────────────┐
  │    TASSEL     │  → View and validate HapMap format
  │               │  → Run GLM/MLM association (optional)
  └───────┬───────┘
          │
          ▼
  ┌───────────────┐
  │   OmicsBox    │  → PCA 
  │               │  → FST 
  └───────┬───────┘
          │ 
          ▼
  ┌───────────────┐
  │  R (custom)   │  → Phenotype cleaning & trait calculation
  │               │  → GLM association: y ~ SNP + breed + location
  │               │  → Genotype & allele frequency
  │               │  → PCA plot (ggplot2)
  └───────┬───────┘
          │
          ▼
      OUTPUTS
  ─────────────────
  • Association results (.csv / .xlsx)
  • PCA plot (.png)
  • Genotype & allele frequency table (.csv / .xlsx)
  • HWE / diversity stats table
```

------------------------------------------------------------------------

# Required R Packages

```{r packages}
# Install if not already installed
pkgs <- c("snpReady", "tidyverse", "ggrepel", "writexl")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
}

library(snpReady)
library(tidyverse)
library(ggrepel)
library(writexl)
```

------------------------------------------------------------------------

# Input Data

## Genotype File (HapMap format)

**File:** `geno_to_assoc.txt`\
**Format:** Tab-delimited text file, HapMap standard\
**Expected structure:**

| Column     | Content                           | Example     |
|------------|-----------------------------------|-------------|
| 1 — rs     | SNP ID                            | rs108968214 |
| 2 — allele | Ref/Alt allele                    | C/G         |
| 3 — chrom  | Chromosome                        | 24          |
| 4 — pos    | Position (bp)                     | 59164671    |
| 5–11       | Metadata (strand, assembly, etc.) | NA          |
| 12 onwards | Animal genotypes (2-char codes)   | CC, CG, GG  |

**Genotype coding:**

| Code              | Meaning      |
|-------------------|--------------|
| CC / GG / TT / AA | Homozygous   |
| CG / AC / GT etc. | Heterozygous |
| NN / --           | Missing      |

> **Note:** Monomorphic SNPs (single allele format in column 2, e.g. `G`
> instead of `C/G`) are removed during QC. These represent SNPs with no
> variation in the population.

------------------------------------------------------------------------

## Phenotype File

**File:** `pheno_to_assoc.txt`\
**Format:** Tab-delimited text file\
**Expected columns:**

| Column | Content | Notes |
|------------------------|------------------------|------------------------|
| animalID | Animal identifier | Must match exactly with HapMap sample IDs |
| breed | Breed name | KK, KKE, Nellore |
| location | Station name | PTH Jelai NS, PTH Tersat Terengganu, etc. |
| BW0 | Birth weight (kg) | True birth weight |
| BW6 | 6-month weight (kg) | 0 = missing → converted to NA |
| BW12 | 12-month / yearling weight (kg) | 0 = missing → converted to NA |
| WW205 | 205-day adjusted weaning weight (kg) | Pre-calculated |
| ADG_0_6 | Pre-weaning ADG (kg/day) | Pre-calculated |
| ADG_6_12 | Post-weaning ADG (kg/day) | Pre-calculated |
| RGR | Relative growth rate (%) | Pre-calculated |

> **Important:** Missing BW6 values are coded as `0` in the raw file —
> these must be converted to `NA` before analysis (handled in Step 2
> below).

------------------------------------------------------------------------

## Growth Trait Formulae

| Trait    | Formula                         | Unit   |
|----------|---------------------------------|--------|
| WW205    | ((BW6 − BW0) / 180) × 205 + BW0 | kg     |
| ADG_0-6  | (BW6 − BW0) / 180               | kg/day |
| ADG_6-12 | (BW12 − BW6) / 180              | kg/day |
| RGR      | ((BW12 − BW0) / BW0) × 100      | \%     |

> Assumed age at 6-month weighing = 180 days (BIF standard for WW205).

***
# Step 1: Genotype QC using snpReady

**Input:** `geno_file` (txt format, transpose geno Animals x SNPs, breed in col2, pre-loaded in RStudio)
**Input:** `pheno_file` (txt format, pre-loaded in RStudio)
**Output:** `geno_coded` matrix (Animals × SNPs, coded 0/1/2)

**Output files:**

| File | Content |
|---|---|
| `Diversity_KK.csv` | Ho, He, MAF, PIC, HWE chiSq, pval for KK breed |
| `Diversity_KKElite.csv` | Same for KK Elite |
| `Diversity_Nellore.csv` | Same for Nellore |

**Column interpretation:**

| Column | Meaning | Flag if... |
|---|---|---|
| MAF | Minor allele frequency | < 0.01 → remove |
| Ho | Observed heterozygosity | Ho << He → possible inbreeding |
| He | Expected heterozygosity | — |
| PIC | Polymorphism info content | < 0.10 → low informativeness |
| chiSq | Chi-square for HWE | — |
| pval | HWE p-value | < 0.05 → deviation from HWE |

## Refer to "zebu_snpReady.R" to run QC & pop analysis

# Step 2: TASSEL — Format Validation

**Tool:** TASSEL 5 (GUI or command line)  
**Input:** `geno_to_assoc.txt` (HapMap)  
**Purpose:** Validate HapMap format, view genotype distribution, optional GLM/MLM run

## How to load in TASSEL

1. Open TASSEL
2. `Data` → `Load` → select `geno_to_assoc.txt`
3. Verify: correct number of SNPs and taxa loaded
4. Check: `Data` → `Genotype Summary` — confirm MAF distribution, missing rate

## Optional: Run GLM/MLM in TASSEL

1. Load phenotype file (tab-delimited, first column = Taxa)
2. `Analysis` → `Association` → `General Linear Model` or `Mixed Linear Model`
3. Select Analysis > Relatedness > Distance Matrix
4. Hover Distance Matrix result > Select Analysis > MDS
5. Hover Genotype + Phenotype + MDS > Select Intersect Join
6. Hover Intersect Join result > Select Analysis > GLM
7. Export results as `.txt`

> **Note:** TASSEL results can be used for comparison with R GLM results.
***

# Step 3: Phenotype Cleaning

**Input:** `pheno_file` (pre-loaded in RStudio)\
**Output:** Cleaned `pheno` data frame

```{r pheno_clean}
pheno <- pheno_file
colnames(pheno)[1] <- "animalID"

# Relabel KKE → KK Elite (or reverse depending on your preference)
pheno$breed[pheno$breed == "KKE"] <- "KK Elite"

# Convert zeros to NA — missing measurements coded as 0 in raw file
for (col in c("BW6", "WW205", "ADG_0_6", "ADG_6_12")) {
  pheno[[col]][pheno[[col]] == 0] <- NA
}
pheno$BW12[pheno$BW12 == 0] <- NA
pheno$RGR[pheno$RGR   == 0] <- NA

# Factor breed and location
pheno$breed    <- factor(pheno$breed, levels = c("KK", "KK Elite", "Nellore"))
pheno$location <- factor(pheno$location)

# Remove specific animals if needed
# pheno <- pheno %>% filter(!animalID %in% c("WKKAM43", "UNAN39", "UNAM55"))

# Check
cat("Animals per breed:\n")
print(table(pheno$breed))
cat("\nValid n per trait:\n")
pheno %>%
  select(BW0, WW205, BW12, ADG_0_6, ADG_6_12, RGR) %>%
  summarise(across(everything(), ~sum(!is.na(.)))) %>%
  print()
```

------------------------------------------------------------------------

```{r impute}
# Impute missing with SNP mean (if any remaining)
cat(sprintf("Missing genotypes: %d\n", sum(is.na(geno_coded))))
geno_coded <- t(apply(geno_coded, 1, function(x) {
  x[is.na(x)] <- round(mean(x, na.rm = TRUE)); x
}))

cat(sprintf("SNPs available: %d\n", nrow(geno_coded)))
```

------------------------------------------------------------------------

# Step 6: GLM Association Analysis

**Model:** `y ~ SNP + breed + location`\
**Input:** `pheno` (cleaned), `geno_coded` (Animals × SNPs) in 0,1,2
formatted by snpReady **Output:** `LM_results.csv`, `LM_results.xlsx`

```{r association}
traits <- c("BW0", "WW205", "BW12", "ADG_0_6", "ADG_6_12", "RGR")

# Match animals
common   <- intersect(pheno$animalID, colnames(geno_coded))
pheno_m  <- pheno %>% filter(animalID %in% common) %>% arrange(animalID)
geno_m   <- geno_coded[, sort(common), drop = FALSE]
stopifnot(all(pheno_m$animalID == colnames(geno_m)))

n_snp   <- nrow(geno_m)
results <- list()

for (trait in traits) {

  y_raw   <- pheno_m[[trait]]
  valid   <- !is.na(y_raw)
  n_valid <- sum(valid)
  cat(sprintf("%-12s  n = %d\n", trait, n_valid))
  if (n_valid < 15) next

  # Scale phenotype to mean=0, sd=1
  y         <- scale(y_raw[valid])[, 1]
  pheno_sub <- pheno_m[valid, ]
  geno_sub  <- geno_m[, valid, drop = FALSE]

  res <- data.frame(
    Trait = trait,
    SNP   = rownames(geno_m),
    beta  = NA_real_, SE = NA_real_,
    t     = NA_real_, p  = NA_real_,
    stringsAsFactors = FALSE
  )

  for (s in seq_len(n_snp)) {
    snp_vec <- as.numeric(geno_sub[s, ])
    if (length(unique(snp_vec[!is.na(snp_vec)])) < 2) next
    fit <- tryCatch(
      lm(y ~ snp_vec + breed + location,
         data = data.frame(y        = y,
                           snp_vec  = snp_vec,
                           breed    = pheno_sub$breed,
                           location = pheno_sub$location)),
      error = function(e) NULL
    )
    if (is.null(fit)) next
    cf <- summary(fit)$coefficients
    if (!"snp_vec" %in% rownames(cf)) next
    res$beta[s] <- round(cf["snp_vec", "Estimate"],   4)
    res$SE[s]   <- round(cf["snp_vec", "Std. Error"], 4)
    res$t[s]    <- round(cf["snp_vec", "t value"],    4)
    res$p[s]    <- round(cf["snp_vec", "Pr(>|t|)"],   6)
  }

  results[[trait]] <- res %>% arrange(p)
}

master <- bind_rows(results)

# Significant hits
cat("\n=== SNPs with p < 0.05 ===\n")
print(master %>% filter(p <= 0.05) %>% arrange(p))

# Save
write.csv(master, "LM_results.csv", row.names = FALSE)
write_xlsx(list("All results" = master,
                "p < 0.05"   = master %>% filter(p <= 0.05)),
           "LM_results.xlsx")
cat("Saved: LM_results.csv, LM_results.xlsx\n")
```

**Result columns:**

| Column | Meaning                                          |
|--------|--------------------------------------------------|
| Trait  | Growth trait name                                |
| SNP    | SNP rs ID                                        |
| beta   | Effect size (in SD units — phenotype was scaled) |
| SE     | Standard error of beta                           |
| t      | t-statistic                                      |
| p      | p-value (nominal)                                |

> **Multiple testing:** FDR and Bonferroni correction are not applied
> here — compute separately using `p.adjust(master$p, method = "BH")`
> and `p.adjust(master$p, method = "bonferroni")`.

**Effect size interpretation (beta in SD units):**

|           | beta     |
|-----------|----------|
| \< 0.2    | Small    |
| 0.2 – 0.5 | Moderate |
| \> 0.5    | Large    |

------------------------------------------------------------------------

# Step 7: Genotype and Allele Frequency

**Input:** `geno_coded` (Animals × SNPs), `pheno` (with breed info)\
**Output:** `Geno_Allele_Frequency.csv`, `Geno_Allele_Frequency.xlsx`

```{r freq}
# Define SNPs of interest — replace with your actual significant SNP IDs
snps_of_interest <- c("SNP1", "SNP2", "SNP3", "SNP4", "SNP5")

# Subset
geno_sub <- t(geno_coded)[, snps_of_interest, drop = FALSE]
# geno_sub is now Animals × SNPs

freq_results <- list()

for (snp in snps_of_interest) {
  for (breed in levels(pheno$breed)) {
    idx    <- pheno$breed == breed
    geno_b <- geno_sub[idx, snp]
    n      <- sum(!is.na(geno_b))
    if (n == 0) next

    n_AA <- sum(geno_b == 0, na.rm = TRUE)
    n_AB <- sum(geno_b == 1, na.rm = TRUE)
    n_BB <- sum(geno_b == 2, na.rm = TRUE)

    freq_results[[paste(snp, breed)]] <- data.frame(
      SNP          = snp,
      Breed        = breed,
      N            = n,
      n_HomRef     = n_AA,
      n_Het        = n_AB,
      n_HomAlt     = n_BB,
      Freq_HomRef  = round(n_AA / n, 4),
      Freq_Het     = round(n_AB / n, 4),
      Freq_HomAlt  = round(n_BB / n, 4),
      Freq_Ref     = round((2*n_AA + n_AB) / (2*n), 4),
      Freq_Alt     = round((2*n_BB + n_AB) / (2*n), 4)
    )
  }
}

freq_table <- bind_rows(freq_results)
print(freq_table)

write.csv(freq_table, "Geno_Allele_Frequency.csv", row.names = FALSE)
write_xlsx(list("Frequency Table" = freq_table),
           "Geno_Allele_Frequency.xlsx")
cat("Saved: Geno_Allele_Frequency.csv\n")
```

**Output columns:**

| Column | Meaning |
|----|----|
| N | Total animals in breed with genotype data |
| n_HomRef | Count of homozygous reference (0) |
| n_Het | Count of heterozygous (1) |
| n_HomAlt | Count of homozygous alternate (2) |
| Freq_HomRef / Het / HomAlt | Genotype frequencies (sum = 1.0) |
| Freq_Ref | Reference allele frequency |
| Freq_Alt | Alternate allele frequency (Freq_Ref + Freq_Alt = 1.0) |

------------------------------------------------------------------------

# Output Summary

| Step | Tool | Output file | Content |
|------------------|------------------|------------------|------------------|
| 4 | snpReady | `Diversity_KK.csv` | Ho, He, MAF, PIC, HWE per SNP — KK |
| 4 | snpReady | `Diversity_KKElite.csv` | Same for KK Elite |
| 4 | snpReady | `Diversity_Nellore.csv` | Same for Nellore |
| 5 | R (ggplot2) | `PCA_breed.png` | PCA plot coloured by breed |
| 6 | R (lm) | `LM_results.csv` | Full GLM association results |
| 6 | R (lm) | `LM_results.xlsx` | GLM results (2 sheets) |
| 7 | R | `Geno_Allele_Frequency.csv` | Genotype & allele freq for significant SNPs |
| 7 | R | `Geno_Allele_Frequency.xlsx` | Same, Excel format |

------------------------------------------------------------------------

# Common Issues and Fixes

| Error | Cause | Fix |
|------------------------|------------------------|------------------------|
| `rownames(geno_sub) != pheno$animalID` | Animal order mismatch | Sort both by animalID before matching |
| `invalid factor level, NA generated` | Editing factor column directly | Convert to character first: `as.character()` then relabel then `factor()` |
| `snp_vec not in rownames(cf)` | SNP monomorphic after subsetting | Skip — no variation to test |
| All results NA | geno orientation wrong (SNPs × Animals instead of Animals × SNPs) | Transpose: `geno_coded <- t(geno_coded)` |
| Zero → not NA | Missing BW6 coded as 0 in raw file | Apply zero-to-NA conversion in Step 2 |

------------------------------------------------------------------------

# Session Info

```{r session}
sessionInfo()
```

------------------------------------------------------------------------

*SOP prepared by ruruadra & claude*\
*Last updated: `r Sys.Date()`*


